---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

The extra packages used in this assignment are:
```{r message=FALSE,warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
```


## Loading and preprocessing the data

In this assignment it is assumed the file 'activity.csv' exists in the working directory. The data is loaded using the function _read.csv()_ and stored as a 'tibbler' in the variable _activity\_data_.
```{r}

## Cheking if data file exists
if(!file.exists("activity.csv")){
    stop("File 'activity.csv' was not found in the working directory.")
}
## Loading data
activity_data <- read.csv(
    "activity.csv",
    header = TRUE,
    colClasses = c("integer","Date","integer")
) %>%
    tbl_df # data stored as in tibbler format

```

The initial data is in a long format with three different variables:

* steps <_int_>: Number of steps taking in a 5-minute interval (missing values are coded as NA)

* date <_date_>: The date on which the measurement was taken in YYYY-MM-DD format. 
The experiment lasted for `r activity_data$date %>% unique %>% length` days, from `r activity_data$date %>% head(1)` to `r activity_data$date %>% tail(1)`.

* interval <_int_>: Identifier for the 5-minute interval in which measurement was taken. 
Per day, there are `r activity_data$interval %>% unique %>% length` such intervals.


```{r}
## A quick look at the data
activity_data
str(activity_data)
summary(activity_data)
```

Having successfully loaded the data, we transform the initial data into a wide format with 62 variables in order to facilitate the analysis of the activity per each day:

* interval ("####") <_date_> : The label of interval on which the number of steps was recorded. The labels are all written with 4 characters, with 0's added on the left when necessary, e.g., "5" -> "0005".
* day ("%Y-%m-%d") <_int_>: The number of steps measured each interval, for all the `r activity_data$date %>% unique %>% length` days.

```{r}
## Spreading the days into columns
activity_data_wide <- activity_data %>% spread(date,steps)

## Adding 0's to the interval label. In mutate() the call to nchar(activity_data_wide$interval[i]) produced an error, thus the use of a for loop.
for(i in 1:dim(activity_data_wide)[1]){
    activity_data_wide$interval[i] <- paste(
        c(
            rep("0",4-nchar(activity_data_wide$interval[i])),
            activity_data_wide$interval[i]),
        collapse = "" 
    )
}

```

In this wide format, each row contains all the information relative to a given interval while each column contains the information for a given day. Quantities per day can be calculated down columns, while quantities for a given interval during the whole period of the experiment can be calculated across rows.

```{r }
## A quick look at the data
activity_data_wide 
```

## What is mean total number of steps taken per day?

To calculate the total number of steps per day, we first disregard the _NA_ values in the daily sum (essentially, this equivalent to replacing _NA_ by 0):
```{r}
## Vector with total number of steps for each day. NA set to 0.
activity_total_nona <- colSums(activity_data_wide[2:62], na.rm = TRUE)
```

We can now plot an histogram of the values of _activity\_total\_nona_ grouped in 25 bins:

```{r}
## Custom function to plot histograms of vectors with the total number of steps per day.
custom_histogram <- function(values, na.rm=TRUE){
    data.frame(steps = values, row.names = NULL) %>% 
    ggplot(aes(steps)) +
    geom_histogram(bins=25, lwd=0.25, color = "darkblue", fill = "lightblue") + 
    geom_vline(aes(xintercept=mean(steps, na.rm = na.rm)), color="darkgreen", linetype="dotdash", size=0.7) +
    geom_vline(aes(xintercept=median(steps, na.rm = na.rm)), color="darkred", linetype="dashed", size=0.7) +
    labs(x="Steps per day",y="Frequency") +
    theme_bw(base_size = 11) +
    theme(
        strip.background = element_blank()
    )
}

## Plotting the histogram
custom_histogram(activity_total_nona)

```

where the mean value, `r mean(activity_total_nona)%>%as.integer`, is marked by a green dot-dashed line and the median, `r median(activity_total_nona)%>%as.integer`, is marked by a red dashed line.

```{r}
## A quick look at the data
summary(activity_total_nona)
```

Notice that in the previous histogram there is a relatively large number of days with very small number of steps (see the leftmost bin in the previous histograms) which seems to be at odds with the rest of the distribution. In order to investigate if this is the result of ignoring the _NA_ values when calculating the number of total steps per day, we first look at the distribution of _NA_ values in the data. If we calculate the relative number of _NA_ values in each day (values between 0 and 1) and display it in a table
```{r}
## Dsitribution of NA values in the data
activity_data_wide[,2:62] %>% apply(
    2,
    FUN = function(x) round(sum(is.na(x))/288,5)
) %>% table
```
we find that there are 8 days where all the intervals have _NA_ values, while in all the other 53 days no _NA_ value is found. This suggests that we should look at the distribution of total steps per day while ignoring the 8 days for which no actual values are available
```{r}
## Vector with total number of steps for each day. Days with NA values have sum = NA.
activity_total_na <- colSums(activity_data_wide[2:62])
```

We can now re-plot an histogram of the values stored in the dataframe _activity\_total\_na_ for different number of bins:
```{r}
## Plotting the histogram
custom_histogram(activity_total_na)
```

This lowers considerably the number of days with 0 total steps as the 8 days with only _NA_ do not appear in the histogram. In fact,  if we now call _summary()_ on _activity\_total\_na_ 
```{r}
summary(activity_total_na)
```
we find that the minimum number of steps in a day is `r min(activity_total_na, na.rm = TRUE)`.

In the following table we compare the values of the median and mean value of steps when (1) ignoring _NA_ values when calculating the total steps per day and (2) ignoring the 8 days where all entries are _NA_ values:

|       |   NA values	|   Median	|   Mean	|
|---	|---    |---	|---	|
|   (1)    |  NA = 0	|   `r median(activity_total_nona, na.rm = TRUE   )%>%as.integer`	|   `r mean(activity_total_nona, na.rm = TRUE)%>%as.integer`	|
|   (2)    | 	NA days removed |   `r median(activity_total_na, na.rm = TRUE   )%>%as.integer`	|   `r mean(activity_total_na, na.rm = TRUE)%>%as.integer`	|

Notice that when the 8 days with _NA_ values are removed (2), the median and mean values are almost identical, in contrast with the situation in (1)

## What is the average daily activity pattern?

Since the _NA_ values in the data occur only for 8 days where no measurements are available for any interval, we can disregard these days while computing the mean number of steps in each interval during the duration of the experiment. 

We store the mean value of steps per interval in the new dataframe _activity\_mean_:
```{r}
## Long dataframe with value of mean number of steps per interval
activity_mean <- activity_data_wide %>%
    transmute(interval, mean = rowMeans(.[2:62],na.rm = TRUE))

```
which has 2 variables:

* interval ("####") <_char_>: The label of interval on which the number of steps was recorded.
* mean <_int_>: The average number of steps for the respective interval over the whole duration of the experiment.

```{r}
## A quick look at the data
activity_mean
```

Below is a time series plot of the evolution of the average number of steps per interval throughout the day
```{r}
## Custom function to plot time series from dataframes like activity_mean
custom_timeseries <- function(df){ 
    df %>% 
    ggplot(aes(as.POSIXct(interval, format = "%H%M"),mean)) + 
    geom_line(color = "darkblue") + 
    scale_x_datetime(date_labels = "%H:%M") +
    labs(x="Time",y="Steps", title = "Average number of steps per 5-minute interval") +
    theme_bw(base_size = 11) +
    theme(
        strip.background = element_blank()
    )
}

## Plotting the time series
custom_timeseries(activity_mean)
```

We store the value of the maximum number of average steps and of the corresponding interval in the vector _vector\_max_: 
```{r}
## Finding the interval with maximum number of average steps
vector_max = filter(activity_mean, mean == max(mean))

```
and report it in the following table

| Interval  | Average steps |
|---      |---      |
| `r vector_max[1,1]` | `r vector_max[1,2]%>%as.integer`|



## Imputing missing values

As stated above, our initial data has 8 days for which all of the 288 intervals have a _NA_ value. Thus, the number of _NA_ values should be `r 8*288`.This  can be check by calling
```{r}
## Number of NA values in the data
sum(is.na(activity_data_wide))
```

The distribution of number of steps and _NA_ values along the day for all 61 days can also be observed in the following heatmap where the _NA_ values are shown  in red
```{r}
## Custom heatmap giving a visual representation of the wide dataframes
custom_heatmap <- function(df) {
    df %>% 
    gather(key = "date", value = "steps",-1) %>%
    ggplot(aes(date,as.POSIXct(interval, format = "%H%M"))) +
    geom_tile(aes(fill = steps), colour = "white") + 
    scale_fill_gradient(low = "white", high = "steelblue", na.value ="red") + 
    scale_y_datetime(date_labels = "%H:%M",expand = c(0, 0)) +
    labs(x="Date",y="Time", title = "Heatmap of the number of steps") +
    theme_bw(base_size = 11) +
    theme(
        strip.background = element_blank(), 
        axis.text.x = element_text(size = 8, angle = 90)
    )
}

## Plotting the heatmap
custom_heatmap(activity_data_wide)
```
The columns (days) in the data frame with _NA_ values can be identified systematically by 
```{r}
## Vector with the colnumber of the columns in the data with NA values
na_cols <- activity_data_wide[2:62] %>% 
    colSums(na.rm=FALSE) %>% 
    is.na %>% 
    which %>% 
    as.integer
```

As a a first strategy fr filling in the missing data, we will use the mean values of number of steps for each 5-minute interval, as calculated in the previous section. We will store the result in the new (288x62) dataframe  _activity\_fill1\_wide_ defined below
```{r}
## Creating new dataframe
activity_fill1_wide <- activity_data_wide

## Replacing NA values with values of activity_mean
for(i in na_cols){
    for(j in 1:dim(activity_fill1_wide)[1]){
        activity_fill1_wide[j,1+i] <- activity_mean[j,2]
    }
}
```

We can now calculate the  number of total steps per day in the new dataframe _activity\_fill1\_wide_
```{r}
## Vector with total number of steps for each day. NA = mean.
activity_total_fill1 <- colSums(activity_fill1_wide[2:62])
```
which we can then use to plot the following histogram
```{r}
## Histogram of total steps per day for new data with NA = mean
custom_histogram(activity_total_fill1)
```

As expected, the substitution of _NA_ values by the mean number of steps in each interval has led to an increase in the number of days with average total number of days, as seen by the increase in frequency of the bin corresponding to the mean and median values. Looking at the summary of _activity\_total\_fill1_ below we find that in this more peaked distribution, the median and mean values almost do not change when compared with the previous results, but the distribution of the quartiles is altere.
```{r}
summary(activity_total_fill1)
```



If we now look at the heatmap of the new dataframe _activity\_fill1\_wide_, we find that the red vertical lines that marked the _NA_ values have been replaced. However, the days with replaced values have timely distributions that standout with respect to the rest of the days, suggesting that this method of filling for missing data is not the most appropriate one. 

```{r}
## Heatmap for new data with NA = mean
custom_heatmap(activity_fill1_wide)
```

As a second strategy to fill in the missing data, we consider the possibility of filling in the _NA_ for a given interval in a given day by [interpolating](https://en.wikipedia.org/wiki/Interpolation) the values of the same interval in the adjacent days. This strategy is somewhat motivated by an apparent correlation between the numbers of steps in the same interval on consecutive days in the heatmaps pictured above. When implementing this strategy we need to account as well for the special cases of the first and last days, which have only one adjacent day to pick values from, and to the pair '2012-11-09' and '2012-11-10' with is the only case of consecutive days with _NA_ values:

* First and last days: for each interval, the new value will be the mean of the same interval for the following or previous two days.
* Days '2012-11-09' and '2012-11-10': for each interval, the new value will be set using a linear interpolation between the corresponding values for the days '2012-11-08' and '2012-11-11'.

We will store the result in the new (288x62) dataframe  _activity\_fill2\_wide_ defined below
```{r}
## Creating new dataframe
activity_fill2_wide <- activity_data_wide

## Replacing NA values
for(i in na_cols){
    ## For i=1 (first day), a weighted average of the values of the second and third days is used giving more weight to the closest column
    if(i==1) {
        for(j in 1:dim(activity_fill2_wide)[1]){
            activity_fill2_wide[j,1+i] <- 
                (2/3)*activity_data_wide[j,1+(i+1)] + (1/3)*activity_data_wide[j,1+(i+2)]
        }
    }
    ## For i=61 (last ay), a weighted average of the values of the penultimate and antepenultimate days is used giving more weight to the closest column
    else if(i==61){
        for(j in 1:dim(activity_fill2_wide)[1]){
            activity_fill2_wide[j,1+i] <- 
                (2/3)*activity_data_wide[j,1+(i-1)] + (1/3)*activity_data_wide[j,1+(i-2)]
        }
    }
    ## If the column i+1 also has a NA value then a weighted average with the values for i-1 and i+2 is used giving more weight to the closest column
    else if(is.na(activity_data_wide[j,1+i+1])) {
        for(j in 1:dim(activity_fill2_wide)[1]){
            activity_fill2_wide[j,1+i] <- 
                (2/3)*activity_data_wide[j,1+i-1] + (1/3)*activity_data_wide[j,1+i+2]
        }
    }
    ## If the column i-1 also has a NA value then a weighted average with the values for i-2 and i+1 is used giving more weight to the closest column
    else if(is.na(activity_data_wide[j,1+i-1])) {
        for(j in 1:dim(activity_fill2_wide)[1]){
            activity_fill2_wide[j,1+i] <- 
                (1/3)*activity_data_wide[j,1+i-2] + (2/3)*activity_data_wide[j,1+i+1]
        }
    }
    ## For all other cases, the mean value of the next-neighbour values is used.
    else {
        for(j in 1:dim(activity_fill2_wide)[1]){
            activity_fill2_wide[j,1+i] <- rowMeans(activity_data_wide[j,1+c(i-1,i+1)])
        }
    }
}

## NOTE: In a more general case, the if/else statements should probably go inside the j-loop. In this case, it seemed more natural to put the j-loop inside the if/else statements which concerned entire columns.
```

We can now calculate the  number of total steps per day in the new dataframe _activity\_fill2\_wide_
```{r}
## Vector with total number of steps for each day. NA = interpolation
activity_total_fill2 <- colSums(activity_fill2_wide[2:62])
```
which we can then use to plot the following histogram
```{r}
## Histogram of total steps per day for new data with NA = interpolation
custom_histogram(activity_total_fill2)
```

From the figure above we find that this second method of filling in for missing data does not lead to such a peaked distribution around the mean value, as in the previous case. In fact, if we look at the summary of _activity\_total\_fill2_ we find that in this case the distribution of the first and fourth quartiles stays more or less the same but the median and mean values are lowered.
```{r}
summary(activity_total_fill2)
```


If we look at the heatmap of the new dataframe _activity\_fill2\_wide_ we find that once more the red vertical lines that marked the _NA_ values have been replaced and that the days with replaced values standout less in the heatmap, although they can still be identified by the _visual smoothing effect_ that they produce. A more complete study could try to understand the interpolation mechanism should use more than the two near-neighbour days and whether the rules used for the first and final day and for the  '2012-11-09'-'2012-11-10' pair could be improved.

```{r}
custom_heatmap(activity_fill2_wide)
```

Finally, in the following table we present the values of the median and mean values of total steps per day when (1) ignoring _NA_ values when calculating the total steps per day; (2) ignoring the 8 days where all entries are _NA_ values; (3) the missing values are replaced by the mean value of the corresponding interval and (4) the missing values are replaced using the simple interpolation method described above.

|       |   NA values	|   Median	|   Mean	|
|---	|---    |---	|---	|
|   (1)    |  NA = 0	|   `r median(activity_total_nona, na.rm = TRUE   )%>%as.integer`	|   `r mean(activity_total_nona, na.rm = TRUE)%>%as.integer`	|
|   (2)    | 	NA days removed |   `r median(activity_total_na, na.rm = TRUE   )%>%as.integer`	|   `r mean(activity_total_na, na.rm = TRUE)%>%as.integer`	|
|   (3)    | 	NA = interval mean |   `r median(activity_total_fill1, na.rm = TRUE   )%>%as.integer`	|   `r mean(activity_total_fill1, na.rm = TRUE)%>%as.integer`	|
|   (4)    | 	NA = linear interpolation |   `r median(activity_total_fill2, na.rm = TRUE   )%>%as.integer`	|   `r mean(activity_total_fill2, na.rm = TRUE)%>%as.integer`	|

## Are there differences in activity patterns between weekdays and weekends?

In order to study the possible difference in patterns between weekdays and weekends, we will use the dataframe _activity\_fill2\_wide_ where the _NA_ values were replaced using the simple interpolation method described above. From that dataframe we create a new data frame _activity\_mean2_ with 3 variables:

* interval ("####") <_char_>: The label of interval on which the number of steps was recorded.
* date <_fct_>: A label that indicates whether the number of steps was averaged over weekends or weekdays.
* mean <_int_>: The average number of steps for the respective interval over the weekend or the weekdays.

```{r }
## Long dataframe with value of mean number of steps per interval
activity_mean2 <- activity_fill2_wide %>%
    gather(key = "date", value = "steps", (2:62)) %>%
    ## Transforming dates in the 'weekend'/'weekday' label
    mutate( date = wday(date) %>% (function(x) ifelse(x %in% c(1,7),"weekend","weekday")) %>% factor) %>% 
    group_by(interval,date) %>%
    summarise(mean = mean(steps))

activity_mean2
```

Using the _custom\_timeseries()_ function defined previously and adding a _facet\_grid()_ that runs over the values of the factor 'date', we can then easily plot the time series of the average number of steps per interval over the weekend and during the weekdays:

```{r}
## Plotting the time series
custom_timeseries(activity_mean2) %>% + facet_grid(activity_mean2$date~.)
```

From the plots above, it is possible to see that, on average, the subject of the experiment started his/hers activity later in the day during the weekends--for weekdays the step number has a sharp increase around 6:00 and another sharp increase around the peak around 8:00 while for weekends the step number increases gradually from 6:00 until a peak around 9:00-- and remained active until later in the day--on weekdays the average number of steps per interval goes below 50 around 19:00 while on weekends it remains above 50 until around 21:00. In addition, there seems to be a higher number of steps between 12:00 and 18:00 on weekends when compared with weekdays.
